---
output:
  pdf_document: default
  html_document: default
  word_document: default
---
# Analyzing the impact of storm events on U.S. population health and economy

## Synopsis:
In examining which types of events are most harmful with respect to population 
health across the US, the analysis found that excessive heat, tornadoes and 
flash floods have had the highest number of fatalities from 1996 to 2011, and 
floods, excessive heat and tornadoes have had the highest number of injuries. 
When examining the impact per occurence of each event, excessive heat still has 
the highest rate of fatalities, but rip current and avalanche rise into the top 
three as well.

Examining which types of events have the greatest economic consequences across
the US, the analysis found floods, hurricanes/typhoons and storm surges/tides
have caused the greatest combined property and crop damage from 1996 to 2011.
Drought meanwhile caused the most crop damage. When examining the impact per 
occurence of each event, floods drop down to fourth, with hurricanes/typhoons 
and storm surges/tides taking 1 and 2 and tropical storms joining the top three.

## Data Processing

Read in data and subset it to pull out state, event type, begin date, 
fatalities, injuries, crop damage and property damage, to make analysis faster.
```{r cache = TRUE}
st <- read.csv("repdata%2Fdata%2FStormData.csv.bz2", stringsAsFactors = FALSE)
st1 <- st[, c(1, 2, 7, 8, 23:28)]
```

Convert BGN_DATE to Date class and subset data after January 1996, because the 
NWS didn't start recording all events until 1996.
```{r}
library(dplyr)
st1$BGN_DATE <- as.Date(st1$BGN_DATE, "%m/%d/%Y %H:%M:%S")
st1 <- st1 %>% mutate(Year = format(st1$BGN_DATE, "%Y")) %>%
        filter(Year >= 1996)
```

Do some initial quick cleaning on the EVTYPE column to get rid of summary data,
make all entries upper case, and remove spaces at the beginning of the entries.

```{r}
st1$EVTYPE <- lapply(st1$EVTYPE, function(v){
        if (is.character(v)) return(toupper(v))
        else return(v)
    })
msk <- grepl("SUMMARY", st1$EVTYPE)
st1 <- st1[!msk, ]
st1$EVTYPE <- gsub("^ ", "", st1$EVTYPE)
```

Next, subset the data further in order to be able to answer the first question:
which types of events are most harmful with respect to population health? Start
by finding which event types had fatalities and subset st1 for only those with 
more than 24 fatalities (the mean is 24.27).

```{r}
deaths <- st1 %>% group_by(EVTYPE) %>% summarise(fatalities = 
            sum(FATALITIES)) %>% filter(fatalities > 24)
q1 <- filter(st1, EVTYPE %in% deaths$EVTYPE)
```

Replace unofficial event types with the official name. The only unofficial event
type unchanged is "Fog", because it could either be Dense Fog or Freezing Fog, 
and there isn't enough information to determine which.

```{r}
q1$EVTYPE <- gsub("EXTREME COLD$", "EXTREME COLD/WIND CHILL", q1$EVTYPE)
q1$EVTYPE <- gsub("HEAVY SURF/HIGH SURF", "HIGH SURF", q1$EVTYPE)
q1$EVTYPE <- gsub("HURRICANE$", "HURRICANE/TYPHOON", q1$EVTYPE)
q1$EVTYPE <- gsub("RIP CURRENTS", "RIP CURRENT", q1$EVTYPE)
q1$EVTYPE <- gsub("WINTER WEATHER/MIX", "WINTER WEATHER", q1$EVTYPE)
q1$EVTYPE <- gsub("TSTM WIND", "THUNDERSTORM WIND", q1$EVTYPE)
q1$EVTYPE <- gsub("LANDSLIDE", "DEBRIS FLOW", q1$EVTYPE)
q1$EVTYPE <- gsub("URBAN/SML STREAM FLD", "FLASH FLOOD", q1$EVTYPE)
```

Then subset and process the data for question two: Which types of events have 
the greatest economic consequences?

First, create new columns with total propdmg and cropdmg using the exponent
column. To do this, create two vectors p and c with the number corresponding to 
the exponent letter in PROPDMGEXP and CROPDMGEXP. Then create the new columns
by multiplying the vectors by the PROPDMG and CROPDMG columns.

```{r}
exp <- function(v) {
    B <- grepl("B", v)
    K <- grepl("K", v)
    M <- grepl("M", v)
    v[B] <- 1000000000
    v[K] <- 1000
    v[M] <- 1000000
    v[!B & !K & !M] <- 1
    as.numeric(v)
}
p <- exp(st1$PROPDMGEXP)
c <- exp(st1$CROPDMGEXP)
st1 <- mutate(st1, TotalPROPDMG = p*PROPDMG, TotalCROPDMG = c*CROPDMG)
```

Find which event types had property and crop damage and subset st1 for only 
those with greater than the mean property or crop damage (1.008e+09 and
9.547e+07, respectively).

```{r}
pdamage <- st1 %>% group_by(EVTYPE) %>% summarise(propdmg = 
                sum(TotalPROPDMG)) %>% filter(propdmg > 1.008e+09)
cdamage <- st1 %>% group_by(EVTYPE) %>% summarise(cropdmg = 
                sum(TotalCROPDMG)) %>% filter(cropdmg > 9.547e+07)
q2 <- filter(st1, EVTYPE %in% pdamage$EVTYPE | EVTYPE %in% cdamage$EVTYPE)
```

Replace unofficial event types with the official name.

```{r}
q2$EVTYPE <- gsub("EXTREME COLD$", "EXTREME COLD/WIND CHILL", q2$EVTYPE)
q2$EVTYPE <- gsub("^FREEZE", "FROST/FREEZE", q2$EVTYPE)
q2$EVTYPE <- gsub("HURRICANE$", "HURRICANE/TYPHOON", q2$EVTYPE)
q2$EVTYPE <- gsub("STORM SURGE$", "STORM SURGE/TIDE", q2$EVTYPE)
q2$EVTYPE <- gsub("WILD/FOREST FIRE", "WILDFIRE", q2$EVTYPE)
q2$EVTYPE <- gsub("TSTM WIND", "THUNDERSTORM WIND", q2$EVTYPE)
```

## Results

### Q1: Across the US, which types of events (EVTYPE) are most harmful with 
### respect to population health?
(Perform this analysis using the q1 data set.)

Excessive Heat, Tornadoes and Flash Floods have the highest number of
fatalities from 1996 to 2011.

```{r}
q1 <- group_by(q1, EVTYPE)
fi <- q1 %>% summarise(Fatalities = sum(FATALITIES), Injuries = 
            sum(INJURIES), Frequency = n()) %>% arrange(desc(Fatalities)) %>%
            rename("EventType" = EVTYPE)

head(fi, 10)
```

Tornadoes, Floods and Excessive Heat meanwhile have the highest number of
injuries from 1996 to 2011.

```{r}
fi <- arrange(fi, desc(Injuries))
head(fi, 10)
```

When examining the impact per occurence of each event, excessive heat still 
has the highest rate of fatalities, at an average of 1.09 per occurence. Other 
events rise into the second and third positions, however, with rip current and 
avalanche coming in second and third with an average 0.79 and 0.59 fatalities 
per occurence, respectively.

```{r}
fi2 <- fi %>% mutate(FatalityRate = Fatalities/Frequency, InjuryRate = 
                         Injuries/Frequency) %>% arrange(desc(FatalityRate))
head(fi2, 10)
```

Looking at injury rates, tsunamis, hurricanes and typhoons, and excessive heat
cause the most injuries per event.

```{r}
fi2 <- arrange(fi2, desc(InjuryRate))
head(fi2, 10)
```

### Q2: Across the US, which types of events have the greatest economic 
### consequences?
(Perform this analysis using q2 data set.)

Floods, hurricanes/typhoons and storm surges/tides caused the greatest combined 
property and crop damage from 1996 to 2011.

```{r}
damage <- q2 %>% group_by(EVTYPE) %>% summarise(PropertyDmg = sum(TotalPROPDMG),
            CropDmg = sum(TotalCROPDMG), Frequency = n()) %>%
            mutate(TotalDamage = CropDmg + PropertyDmg) %>%
            arrange(desc(TotalDamage)) %>% rename("EventType" = EVTYPE)
head(damage)
```

Floods, hurricanes/typhoons and storm surges/tides have caused the greatest
property damage from 1996 to 2011.

```{r}
damageP <- arrange(damage, desc(PropertyDmg))
head(damageP, 10)
ptop5 <- damageP[1:5, ]
library(ggplot2)
g <- ggplot(ptop5, aes(x = EventType, y = PropertyDmg))
g + geom_bar(stat = "identity", fill = "#99d8c9") + labs(x = "Event Type", 
                            y = "Property Damage (in dollars)") +
    theme(axis.text.x = element_text(angle = 15, size = 7)) +
    ggtitle('Top 5 Storm Causes of Property Damage')
```

Drought has meanwhile caused the most crop damage, with hurricanes/typhoons
and floods still in the top three.

```{r}
damageC <- arrange(damage, desc(CropDmg))
head(damageC, 10)
ctop5 <- damageC[1:5, ]
h <- ggplot(ctop5, aes(x = EventType, y = CropDmg))
h + geom_bar(stat = "identity", fill = "steel blue") + labs(x = "Event Type", 
    y = "Crop Damage (in dollars)") +
    theme(axis.text.x = element_text(angle = 15, size = 7)) +
    ggtitle('Top 5 Storm Causes of Crop Damage')
```

When examining the economic impact per occurence of each event, floods drop down
to fourth, with hurricanes/typhoons and storm surges/tides taking 1 and 2. 
Tropical storms rise to the third ranking.

```{r}
damage <- damage %>% mutate(TotalDmgRate = TotalDamage/Frequency) %>% 
                        arrange(desc(TotalDmgRate))
head(damage)
```